### 2015-03-14 theory.tex
update background on lockset analysis
update on theory section

To counter these limitations, we apply the idea of Eraser's lockset analysis in a static context.

We propose \emph{static pair-wise lockset analysis}, a novel verification technique for proving data race freedom in device drivers. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

@@

We now formalise our approach to verifying data race freedom in a device driver by statically analysing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all device drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.

For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$. Furthermore, for each entry point $\mathit{ep}_{i}$, we compute $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which maps each $m$ to true if and only if $\mathit{ep}_{i}$ writes to $m$ during some execution.

\begin{theorem}
\label{theorem:locksets}
For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the device driver with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
\end{theorem}

A sketch of how the above theorem can be proved is as follows. Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$. By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$. By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both entry points $ep_1$ and $ep_2$. As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

@@

Our approach begins by performing \emph{two-thread reduction}, an abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent driver, and then performs \emph{pair-wise sequentialisation}, which combines the two arbitrary threads in a single sequential pair. This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point. This process repeats until all pairs of entry points have been sequentialised. To achieve soundness, each time an entry point performs a read access to a shared resource, we return a nondeterministic value. This over-approximates any effects from all the unmodeled threads on the device driver shared state.

>>>>

We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.

@@

We now formalise our approach to verifying data race freedom in a kernel module by statically analysing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a module, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this module. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all modules that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking. The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.

For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$. Furthermore, for each entry point $\mathit{ep}_{i}$, we compute its read-set $R_{i}\lbrack m\rbrack \rightarrow \{true, false\}$ and its write-set $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which map each $m$ to true if and only if $\mathit{ep}_{i}$ reads or writes respectively to $m$ during some execution. In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location. \PDComment{any concrete reference that quantifiers might be harder for SMT solving?}

\begin{theorem}
\label{theorem:locksets}
For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the kernel module with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
\end{theorem}

A sketch of how the above theorem can be proved is as follows. Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$. By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$. By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both $ep_1$ and $ep_2$. As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

@@

Our approach begins by performing \emph{two-thread reduction}, an abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent driver, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair. This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point. This process repeats until all pairs of entry points have been sequentialised. To achieve soundness, each time an entry point performs a read access to a shared resource, we return a nondeterministic value. This over-approximates any effects from all the unmodeled threads on the device driver shared state.
